Genetic signatures of evolutionary rescue by a selective sweep

### 2020-05-10 manuscript/main.tex
second round of revisions, 1/k factor in theta added, moved much to sup mat

One of the most useful models in population genetics is that of a selective sweep and the consequent hitch-hiking of linked neutral alleles.
While variations on this model typically assume constant population size, many instances of strong selection and rapid adaptation in nature may co-occur with complex demography.
Here we extend the hitch-hiking model to evolutionary rescue, where adaptation and demography not only co-occur but are intimately entwined. 
Our results show how this feedback between demography and evolution determines -- and restricts -- the genetic signatures of evolutionary rescue, and how these differ from the signatures of sweeps in populations of constant size.
In particular, we find rescue to harden sweeps from standing variance or new mutation (but not from migration), reduce nucleotide diversity both at the selected site and genome-wide, and increase the range of observed Tajima's $D$ values.
For a given initial rate of population decline, the feedback between demography and evolution makes all of these differences more dramatic under weaker selection, where bottlenecks are prolonged. 
Nevertheless, it is likely difficult to infer the co-incident timing of the sweep and bottleneck from these simple signatures, never-mind a feedback between them. 
Temporal samples spanning contemporary rescue events may offer one way forward.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Methods}
\label{sec:methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Consider a population of size $N(t)$ with a beneficial allele, $A$, at frequency $p(t)$ and an ancestral allele, $a$, at frequency $q(t)=1-p(t)$.
Assume non-overlapping generations and let $W_{AA}$, $W_{Aa}$, and $W_{aa}$ be the absolute fitness (expected number of offspring) of each genotype.
Then with random mating the expected change in allele frequency \citep[equation 5.2.13 in][]{crow1970introduction} and population size in one generation is
\begin{equation}\label{eq:deltapdeltaN}
\begin{aligned}
\Delta p(t) &\equiv p(t+1) - p(t) = p(t) q(t) \frac{  (W_{AA} - W_{Aa} )p(t) +  (W_{Aa} - W_{aa})q(t)}{\overline{W}(t)}  \\
\Delta N(t) &\equiv N(t+1) - N(t) = (\overline{W}(t) - 1) N(t),
\end{aligned}
\end{equation}
with $\overline{W}(t)$ the population mean fitness.

Here we are interested in the scenario where a population composed of primarily $aa$ genotypes is declining at some rate $d$, i.e., $W_{aa} = 1-d$.
The beneficial allele, $A$, is then assumed to act multiplicatively with the fitness of the ancestral background, such that $W_{Aa} = (1-d)(1+ h s)$ and $W_{AA} = (1-d)(1+s)$.
Throughout the text we assume additivity at the selected locus, $h=1/2$, such that the allele frequency \citep[c.f., equation 5.3.12 in][]{crow1970introduction} and population dynamics can be approximated by
\begin{equation}\label{eq:ptnt}
\begin{aligned}
p(t) &\approx \left( 1 + \frac{1-p(0)}{p(0)} e^{-s t/2} \right)^{-1}\\
N(t) &\approx N(0) e^{ \left[ (1+s)(1-d)-1\right]t} \left(\frac{p(0)}{p(t)} \right)^{2(1-d)}.
\end{aligned}
\end{equation}

@@

Equations \ref{eq:deltapdeltaN} and \ref{eq:ptnt} consider only the deterministic dynamics. 
We are, however, primarily concerned with a stochastic event -- the establishment of a (potentially rare) beneficial allele.
In rescue we are particularly interested in only those instances where the beneficial allele establishes (otherwise the population is extinct), which creates a bias away from the deterministic trajectory.
As shown in \cite{smith1971use} \citep[see][for an application to evolutionary rescue]{orr2014population}, we can approximate this bias by assuming that a single copy of an allele that establishes with probability $P_\mathrm{est}$ instantaneously increases from a frequency of $1/(2N)$ to $1/(2NP_\mathrm{est})$.
In essence, dividing the initial allele frequency by the probability of establishment implies that an allele escaping random loss when rare will quickly increase to a larger frequency than expected before settling into its deterministic trajectory.
As we will see below, we can combine this logic with the waiting time distribution for an establishing mutation to arrive to derive an effective initial allele frequency, $p_0$.
Using the deterministic equations (Equation \ref{eq:deltapdeltaN} or \ref{eq:ptnt}) with $p(0) = p_0$ then approximates the forward-time trajectory of an allele that is destined to fixation.

We derive the probability of establishment, $P_\mathrm{est}$, assuming alleles do not interact (i.e., a branching process).
Under this assumption a single copy of an allele that is expected to leave $1+\epsilon$ copies in the next generation with a variance of $v$ has an establishment probability of \citep[][p.\ 172, see also \citealp{feller1951diffusion}, equation 5.7, for a derivation from a diffusion approximation]{allen2010introduction}
\begin{equation}\label{eq:pest}
P_\mathrm{est} = 1 - e^{-2 \epsilon/v}.
\end{equation}
For example, in a Wright-Fisher population of size $N$ a rare allele with selection advantage $s$ is expected to leave $1+s$ copies, with a variance of $1+s+O(1/N)$.
Thus $\epsilon=s$ and, in a large population with weak selection, $v \approx 1$, such that the probability of establishment is roughly $1-e^{2s}$ \citep[][p.\ 80]{fisher1999genetical} or nearly $2s$ \citep{haldane1927mathematical}. 
In our case, in a declining population ($d>0$) we have $\epsilon = W_{Aa} - 1 = (1 + s h)(1-d)-1 \approx s h - d$ and $v$ depends on the particular choice of lifecycle (e.g., see \nameref{sec:pestNe}).
When $h=1$ and $\sigma^2=1$, we recover the probability of establishment of a weakly beneficial mutation in an exponentially growing or declining haploid population with a Poisson offspring distribution, $P_\mathrm{est} \approx 2(s - d)$ \citep{otto1997probability}.
In a population at carrying capacity (e.g., with $d\geq0$) the growth rate is instead $\epsilon = W_{Aa}/W_{aa} - 1 = (1 + s h)-1 = s h$.

@@

Above we have stitched together the initial stochastic phase of the allele frequency and population dynamics with the deterministic phase by using $p(0)=p_0$. 
We next incorporate the final phase, fixation, which is also stochastic.
Incorporating this final phase is important here \citep[as opposed to studies that predict only the forward-time dynamics; e.g.,][]{orr2014population} because in looking at genetic signatures we are primarily concerned with the dynamics looking backwards in time from the point of fixation, and this third phase determines how fast fixation occurs.
As shown in \cite{martin2015simple}, we can treat the backward-in-time dynamics from fixation as we did the forward-in-time dynamics at establishment. 
In particular, the probability the heterozygote establishes (backward in time) in a population of mutant homozygotes is $P_\mathrm{est}$ (Equation \ref{eq:pest}) with the growth rate of a rare heterozygote in this population, $\epsilon$, replaced by $-\epsilon$ (let's call this probability $P_\mathrm{est}^-$).
For instance, if the fitness of the heterozygote is $W_{Aa} = (1-d)(1+ s h)$ and the fitness of the mutant homozygote is $W_{AA} = (1-d)(1+s)$ and the population is at carrying capacity then $-\epsilon = 1 - W_{Aa}/W_{AA}=1 - (1+ s h)/(1+ s) \approx s(1-h)$.
This does not depend on demography, $d$, in contrast to the growth rate of the heterozygote at establishment (see \nameref{sec:initial_frequency}).
Note that the variance in the number of offspring, $v$, may also be different at the time of fixation than it was at the time of establishment (see \nameref{sec:pestNe}).
Given the heterozygote establishes it will quickly increase to $1/P_\mathrm{est}^-$ copies, implying that fixation effectively occurs when the deterministic trajectory reaches a frequency of $p_f = 1-1/(2 N P_\mathrm{est}^-)$.
Below we will explore scenarios where fixation is expected to occur while the population is at carrying capacity. 
Thus in all cases we have the same effective final allele frequency,
\begin{equation}\label{eq:pf}
p_f = 1 - \frac{1}{2 N(0) P_\mathrm{est}},
\end{equation}
where $P_\mathrm{est}$ is given by Equation \ref{eq:pest} with $\epsilon = 1 - (1 + s h)/(1 +  s)$ and $v$ as given in \nameref{sec:pestNe}.

@@

Setting $t = t_f - \tau$ in Equation \ref{eq:ptnt} gives an approximation of the dynamics of allele frequency and population size backward in time,
\begin{equation}\begin{aligned}\label{eq:backwardsptnt}
p^\prime(\tau) &\approx \left(1 + \frac{1 - p_f}{p_f} e^{s \tau/2} \right)^{-1}\\
N^\prime(\tau) &\approx N(0) e^{-d (t_f-\tau)} \left(\frac{1-p_0}{1-p^\prime(\tau)}\right)^{2(1-d)},
\end{aligned}\end{equation}
from the point of fixation, $\tau=0$, to the time of establishment, $\tau=t_f$.
Note that the backward-time allele frequency dynamics do not depend on the initial frequency or $d$ except in determining the maximum value of $\tau$.
Thus the backward-time allele frequency dynamics are expected to follow the same trajectory in rescue as in populations of constant size with the same carrying capacity (such that the final frequency, $p_f$, is the same).
Meanwhile, the backward-time population size dynamics depend heavily on the initial frequency as the timing of the sweep determines the depth and duration of the bottleneck.

@@

To explore the genetic patterns created by evolutionary rescue we next consider a random sample of chromosomes at the time the beneficial allele fixes.
Focusing on a neutral locus that is recombination distance $r$ from the selected site, we are interested in calculating the rate of coalescence, the rates of recombination and mutation off the selected background, and the rate of migration out of the population.
If our sample of alleles has $k$ distinct ancestors on the selected background $\tau$ generations before fixation, these rates are approximately (table 1 in \citealp{hudson1988coalescent}; equation 16 in \citealp{pennings2006soft2}; see \nameref{sec:derive_coalescent} for derivations)
\begin{equation}\label{eq:coal_rates}
\begin{aligned}
p_{\mathrm{coal}}(k,\tau) &= \binom{k}{2} \frac{1}{2N_e'(\tau) p^\prime(\tau)}\\
p_{\mathrm{rec}}(k,\tau) &= k r [1 - p^\prime(\tau)]\\
p_{\mathrm{mut}}(k,\tau) &= k \frac{u [1 - p^\prime(\tau)]}{p^\prime(\tau)}\\
p_{\mathrm{mig}}(k,\tau) &= k \frac{m}{2N^\prime(\tau)p^\prime(\tau)},
\end{aligned}
\end{equation} 
where $N_e'(\tau)$ is the effective population size $\tau$ generations before fixation.

With slow changes in population size, $N(t-1)\approx N(t)$ and the mean number of gametes contributed to the next generation by each diploid individual in the current generation is $2$.
The inbreeding and variance effective population size, $N_e(t)$, is then roughly $(4N(t) - 2)/(\sigma^2+2)$ \citep[equation 7.6.4.3 in][]{crow1970introduction}, where $\sigma^2$ is the variance in the per capita number of gametes contributed to the next generation.
Therefore, in a large population, $N(t)\gg1$, the ratio of the effective size to the census size is roughly $N_e(t)/N(t) \approx 4/(2+\sigma^2)$, where $\sigma^2$ depends on the particular lifecycle (e.g., see \nameref{sec:pestNe}).

@@

We now use the instantaneous event rates (Equation \ref{eq:coal_rates}) to calculate the probability that the most recent event is $\tau$ generations before fixation and is either coalescence, recombination, mutation, or migration. 
Letting $i,j\in\{\mathrm{coal, rec, mut, mig}\}$, the probability that $i$ is the most recent event, and occurs $\tau$ generations before fixation, is \citep[c.f., equation 6 in][]{pennings2006soft}
\begin{equation}\label{eq:Pikt}
\begin{aligned}
P_i(k,\tau) &= p_{i}(k,\tau) \left( \prod_{j\neq i} [1 - p_{j}(k,\tau)] \right) \prod_{l=0}^{\tau-1} \left( \prod_{j} [1 - p_{j}(k,l)] \right)\\
&\approx p_{i}(k,\tau) \exp \left( -\sum_{j} \int_{0}^{\tau} p_{j}(k,l) \mathrm{d}l \right),
\end{aligned}
\end{equation}
i.e., the waiting time for an inhomogeneous exponential random variable.
The approximation assumes the $p_i(k,\tau)$ are small enough such that at most one event happens each generation, with small probability, and the changes in the $p_i(k,\tau)$ from one generation to the next are small enough that we can approximate a sum across $\tau$ with an integral.
As a technical aside, to speed computation we analytically solve the integrals and evaluate Equation \ref{eq:Pikt} under the assumption of weak selection and exponential population growth (Equation \ref{eq:ptnt}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Genetic diversity}
\label{sec:rescue_heterozygosity}

One classic pattern of genetic variation produced by a selective sweep is a dip in genetic diversity around the selected site \citep{smith1974hitch,kaplan1989hitchhiking}. 
Here we consider the average number of nucleotide differences between two randomly sampled sequences, $\pi$ \citep{tajima1983evolutionary}, focusing on sequences of length 1 (i.e., heterozygosity).

We first consider our expectation for $\pi$ at a site that is far enough away from the selected site to be unaffected by hitchhiking, which will provide us with an expectation for  genome-wide average diversity.
While not directly impacted by the sweep, these loci are indirectly impacted as the sweep dictates the severity of the population bottleneck.
Our expectation for $\pi$ at such a site in a population of constant effective size, $N_e$, is simply $\theta = 4 N_e U$ \citep{watterson1975number}, with $U$ the per base per generation mutation rate at neutral loci.
Ignoring neutral mutation input during the sweep (which will cause us to underestimate diversity), the $\pi$ at a sufficiently loosely linked site is this neutral expectation times the probability a sample of size two does not coalesce during the period of interest (in our case, from the time we take the sample at the time of fixation, $\tau=0$, until the time of environmental change, $\tau=t_f$).
This is (c.f., equation 4 in \citealp{slatkin1991pairwise} and equation 7 in \citealp{griffiths1994sampling})
\begin{equation}\label{eq:EH_rescue_loose}
\mathbb{E}[\pi | \mathrm{unlinked}]\approx \theta \exp \left( - \int_0^{t_f} \frac{1}{2N_e'(\tau)} \mathrm{d}\tau \right). 
\end{equation}
In a population of constant size we simply use $\mathbb{E}[\pi | \mathrm{unlinked}]=\theta$, which is more accurate as it incorporates neutral mutations occurring during the sweep.

We next consider sites that are closer to the selected locus, such that they are directly affected by the selective sweep through hitchhiking.
To keep the analysis simple, we assume that if recombination or mutation moves one of the sampled alleles to the ancestral background before the two samples coalesce then it is as if both samples were on the ancestral background from the start and therefore coalesce with each other as if they were at an unlinked locus (Equation \ref{eq:EH_rescue_loose}).
This assumption therefore ignores the time it takes for the two samples to arrive on the ancestral background and is therefore expected to overestimate coalescence rates.
There is also the possibility of no events occurring in the history of the sample during the sweep.
Then, if the sweep arose from mutation or migration or from a single copy of the beneficial allele ($\kappa=1$), the sample must coalesce (i.e., we observe a hard sweep). 
If we instead start with more than one beneficial copy, $\kappa>1$, then under the assumption that each copy of the beneficial allele in the standing variance has a recent and unique mutational origin \citep[which we will assume throughout for simplicity, noting this effectively implies the beneficial allele was sufficiently deleterious before the environmental change; c.f.,][]{prezeworski2005signature}, the samples are independent draws from a neutral population (i.e., the samples to do not coalesce and we observe a soft sweep).
Ignoring neutral mutations during the sweep, a simple approximation for diversity at any location in the genome is then
\begin{equation}\label{eq:EH_rescue}
\begin{aligned}
\mathbb{E}[\pi ] \approx P_\mathrm{off}(2,t_f) \mathbb{E}[\pi | \mathrm{unlinked}] + P_\emptyset(2,t_f) \delta_{\kappa>1} \theta,
\end{aligned}
\end{equation}
where $P_\mathrm{off}(k,t_f) = \int_0^{t_f} [P_\mathrm{rec}(k,\tau) + P_\mathrm{mut}(k,\tau)] \mathrm{d}\tau$ is the probability of recombination or mutation before coalescence during the sweep, $P_\emptyset(k, t_f) = \exp \left( -\sum_{j} \int_{0}^{t_f} p_{j}(k,\tau) \mathrm{d}\tau \right)$ is the probability that no events have occurred in the history of the sample during the sweep ($j\in\{\mathrm{rec},\mathrm{mut},\mathrm{coal}\}$), and $\delta_{\kappa>1}$ is 1 if the sweep arose from more than one copy of the beneficial allele ($\kappa>1$) and 0 otherwise.
Equation \ref{eq:EH_rescue} is a function of recombination distance, $r$, and approaches $\mathbb{E}[\pi | \mathrm{unlinked}]$ as $r$ gets large.

@@

Code used to derive and plot all results (Python scripts and Mathematica notebook) is available at \url{https://github.com/mmosmond/rescueCoalescent.git}.
File S1 refers to the Mathematica notebook, which is also provided in freely accessible forms (CDF and PDF).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{A simultaneous, but independent, sweep and bottleneck}

>>>>

One of the most useful models in population genetics is that of a selective sweep and the consequent hitch-hiking of linked neutral alleles.
While variations on this model typically assume constant population size, many instances of strong selection and rapid adaptation in nature may co-occur with complex demography.
Here we extend the hitch-hiking model to evolutionary rescue, where adaptation and demography not only co-occur but are intimately entwined. 
Our results show how this feedback between demography and evolution determines -- and restricts -- the genetic signatures of evolutionary rescue, and how these differ from the signatures of sweeps in populations of constant size.
In particular, we find rescue to harden sweeps from standing variance or new mutation (but not from migration), reduce genetic diversity both at the selected site and genome-wide, and increase the range of observed Tajima's $D$ values.
For a given initial rate of population decline, the feedback between demography and evolution makes all of these differences more dramatic under weaker selection, where bottlenecks are prolonged. 
Nevertheless, it is likely difficult to infer the co-incident timing of the sweep and bottleneck from these simple signatures, never-mind a feedback between them. 
Temporal samples spanning contemporary rescue events may offer one way forward.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Materials and Methods}
\label{sec:methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Consider a population of size $N(t)$ with a beneficial allele, $A$, at frequency $p(t)$ and an ancestral allele, $a$, at frequency $1-p(t)$.
Assume non-overlapping generations and let $W_{AA}$, $W_{Aa}$, and $W_{aa}$ be the absolute fitness (expected number of offspring) of each genotype.
%Then with random mating the expected change in allele frequency \citep[equation 5.2.13 in][]{crow1970introduction} and population size in one generation is
%\begin{equation}\label{eq:deltapdeltaN}
%\begin{aligned}
%\Delta p(t) &\equiv p(t+1) - p(t) = p(t) q(t) \frac{  (W_{AA} - W_{Aa} )p(t) +  (W_{Aa} - W_{aa})q(t)}{\overline{W}(t)}  \\
%\Delta N(t) &\equiv N(t+1) - N(t) = (\overline{W}(t) - 1) N(t),
%\end{aligned}
%\end{equation}
%with $\overline{W}(t)$ the population mean fitness.
%\textcolor{red}{no longer use this equation so could move to sup mat}
%
We are interested in the scenario where a population composed of primarily $aa$ genotypes is declining at some rate $d$, i.e., $W_{aa} = 1-d$.
The beneficial allele, $A$, is then assumed to act multiplicatively with the fitness of the ancestral background, such that $W_{Aa} = (1-d)(1+ h s)$ and $W_{AA} = (1-d)(1+s)$.
Throughout the text we assume random mating, weak selection, $s<<1$, and additivity at the selected locus, $h=1/2$, such that the allele frequency \citep[c.f., equation 5.3.12 in][]{crow1970introduction} and population dynamics ($N(t+1) = \overline{W}(t) N(t)$, with $\overline{W}(t)$ the population mean fitness) can be approximated by
\begin{equation}\label{eq:ptnt}
\begin{aligned}
p(t) &\approx \left( 1 + \frac{1-p(0)}{p(0)} e^{-s t/2} \right)^{-1}\\
N(t) &\approx N(0) e^{ \left[ (1+s)(1-d)-1\right]t} \left(\frac{p(0)}{p(t)} \right)^{2(1-d)}.
\end{aligned}
\end{equation}
Equation \ref{eq:ptnt} allows for unbounded population growth; to prevent this we impose a simple carrying capacity at $N(0)$ by setting $N(t)$ to be the minimum of $N(0)$ and the value given in Equation \ref{eq:ptnt}.

@@

Equation \ref{eq:ptnt} considers only the deterministic dynamics. 
We are, however, primarily concerned with a stochastic event: the establishment of a (potentially rare) beneficial allele.
In rescue we are particularly interested in only those instances where the beneficial allele establishes -- otherwise the population is extinct.
Conditioning on establishment creates a bias away from the deterministic trajectory.
As shown in \cite{smith1971use} \citep[see][for an application to evolutionary rescue]{orr2014population}, we can approximate this bias by assuming that a single copy of an allele that establishes with probability $P_\mathrm{est}$ instantaneously increases from an initial frequency of $1/(2N)$ to $1/(2NP_\mathrm{est})$.
In essence, dividing the initial allele frequency by the probability of establishment implies that an allele escaping random loss when rare will quickly increase to a larger frequency than otherwise expected before settling into its deterministic trajectory.
As we will see below, we can combine this logic with the waiting time distribution for an establishing mutation to arrive to derive an effective initial allele frequency, $p_0$.
Using the deterministic prediction (Equation \ref{eq:ptnt}) with $p(0) = p_0$ then approximates the forward-time trajectory of an allele that is destined to fixation.

We derive the probability of establishment, $P_\mathrm{est}$, assuming alleles do not interact (i.e., a branching process).
Under this assumption a single copy of an allele that is expected to leave $1+\epsilon$ copies in the next generation with a variance of $v$ has an establishment probability of \citep[][p.\ 172, see also \citealp{feller1951diffusion}, equation 5.7, for a derivation from a diffusion approximation]{allen2010introduction}
\begin{equation}\label{eq:pest}
P_\mathrm{est} = 1 - e^{-2 \epsilon/v}.
\end{equation}
%For example, in a Wright-Fisher population of size $N$ a rare allele with selection advantage $s$ is expected to leave $1+s$ copies, with a variance of $1+s+O(1/N)$.
%Thus $\epsilon=s$ and, in a large population with weak selection, $v \approx 1$, such that the probability of establishment is roughly $1-e^{2s}$ \citep[][p.\ 80]{fisher1999genetical} or nearly $2s$ \citep{haldane1927mathematical}. 
When $d\leq0$ we ignore density-dependence and the growth rate of the establishing heterozygote then depends only on its absolute fitness, $\epsilon = W_{Aa} - 1 = (1 + s h)(1-d)-1 \approx s h - d$.
The variance, $v$, depends on particulars of the lifecycle (see \nameref{sec:pestNe} for details).
When $h=1$ and $v=1$ we retrieve the probability of establishment of a weakly beneficial mutation in an exponentially growing or declining haploid population with a Poisson offspring distribution, $P_\mathrm{est} \approx 2(s - d)$ \citep{otto1997probability}.
%When $d\geq0$ the population remains at carrying capacity and the growth rate of the establishing heterozygote only depends on its relative fitness, $\epsilon = W_{Aa}/W_{aa} - 1 = (1 + s h)-1 = s h$, which is independent of demography, $d$.

@@

Above we have stitched together the initial stochastic phase of the allele frequency and population dynamics with the deterministic phase by using $p(0)=p_0$. 
We next incorporate the final phase, fixation, which is also stochastic.
Incorporating this final phase is important here \citep[as opposed to studies that predict only the forward-time dynamics; e.g.,][]{orr2014population} because in looking at genetic signatures we are primarily concerned with the dynamics backwards in time from the point of fixation, and this third phase determines how fast fixation occurs.

As shown in \cite{martin2015simple}, we can treat the backward-in-time dynamics from fixation as we did the forward-in-time dynamics at establishment. 
In particular, the probability the heterozygote establishes (backward in time) in a population of mutant homozygotes is also given by Equation \ref{eq:pest} but with the growth rate of a rare heterozygote in this population, $\epsilon$, replaced by $-\epsilon$ (let's call this probability $P_\mathrm{est}^-$).
For instance, if the fitness of the heterozygote is $W_{Aa} = (1-d)(1+ s h)$ and the fitness of the mutant homozygote is $W_{AA} = (1-d)(1+s)$ and the population is at carrying capacity then the growth rate of the heterozygote depends only on its relative fitness, $-\epsilon = 1 - W_{Aa}/W_{AA} = 1 - (1+ s h)/(1+ s) \approx s(1-h)$, which is independent of demography, $d$.
%Note that $\epsilon$ does not depend on demography, $d$, in contrast to the growth rate of the heterozygote at establishment in a declining population (see \nameref{sec:initial_frequency}).
%The variance in the number of offspring, $v$, may also be different at the time of fixation than it was at the time of establishment (see \nameref{sec:pestNe}).
%\textcolor{red}{move to sup mat?}

Given the heterozygote establishes (backward-in-time) it will quickly increase to $1/P_\mathrm{est}^-$ copies, implying that in a population of size $N$ fixation effectively occurs when the deterministic trajectory reaches a frequency of $p_f = 1-1/(2 N P_\mathrm{est}^-)$.
Below we will explore scenarios where fixation is expected to occur while the population is at carrying capacity, $N(0)$. 
Thus in all cases we have the same effective final allele frequency,
\begin{equation}\label{eq:pf}
p_f = 1 - \frac{1}{2 N(0) P_\mathrm{est}},
\end{equation}
where $P_\mathrm{est}$ is given by Equation \ref{eq:pest} with $\epsilon = 1 - (1 + s h)/(1 +  s)$ and $v$ as given in \nameref{sec:pestNe}.

@@

Setting $t = t_f - \tau$ in Equation \ref{eq:ptnt} gives an approximation of the dynamics of allele frequency and population size backward in time,
\begin{equation}\begin{aligned}\label{eq:backwardsptnt}
p^\prime(\tau) &\approx \left(1 + \frac{1 - p_f}{p_f} e^{s \tau/2} \right)^{-1}\\
N^\prime(\tau) &\approx N(0) e^{-d (t_f-\tau)} \left(\frac{1-p_0}{1-p^\prime(\tau)}\right)^{2(1-d)},
\end{aligned}\end{equation}
from the point of fixation, $\tau=0$, to the time of establishment, $\tau=t_f$.
Note that the backward-time allele frequency dynamics do not depend on the initial frequency, $p_0$, or demography, $d$, except in determining the maximum value of $\tau$, $t_f$.
Thus the backward-time allele frequency dynamics are expected to be the same in rescue as in populations of constant size with the same carrying capacity (such that the final frequency, $p_f$, is the same).
Meanwhile, the backward-time population size dynamics depend heavily on the initial frequency as the time at which the sweep establishes determines the depth and duration of the bottleneck.

@@

To explore the genetic patterns created by evolutionary rescue we next consider a random sample of chromosomes at the time the beneficial allele fixes.
Focusing on a neutral locus that is recombination distance $r$ from the selected site, we are interested in calculating the rate of coalescence, the rates of recombination and mutation off the selected background, and the rate of migration out of the population.
If our sample of alleles has $k$ distinct ancestors on the selected background $\tau$ generations before fixation, these rates are approximately (table 1 in \citealp{hudson1988coalescent}; equation 16 in \citealp{pennings2006soft2}; see \nameref{sec:derive_coalescent} for derivations)
\begin{equation}\label{eq:coal_rates}
\begin{aligned}
p_{\mathrm{coal}}(k,\tau) &= \binom{k}{2} \frac{1}{2N_e'(\tau) p^\prime(\tau)}\\
p_{\mathrm{rec}}(k,\tau) &= k r [1 - p^\prime(\tau)]\\
p_{\mathrm{mut}}(k,\tau) &= k \frac{u [1 - p^\prime(\tau)]}{p^\prime(\tau)}\\
p_{\mathrm{mig}}(k,\tau) &= k \frac{m}{2N^\prime(\tau)p^\prime(\tau)},
\end{aligned}
\end{equation} 
where $N_e'(\tau)$ is the effective population size $\tau$ generations before fixation (see \nameref{sec:pestNe} for details).

@@

We now use the instantaneous event rates (Equation \ref{eq:coal_rates}) to calculate the probability that the most recent event is $\tau$ generations before fixation and is either coalescence, recombination, mutation, or migration. 
Letting $i,j\in\{\mathrm{coal, rec, mut, mig}\}$, the probability that $i$ is the most recent event and occurs $\tau$ generations before fixation is \citep[c.f., equation 6 in][]{pennings2006soft}
\begin{equation}\label{eq:Pikt}
\begin{aligned}
P_i(k,\tau) &= p_{i}(k,\tau) \left( \prod_{j\neq i} [1 - p_{j}(k,\tau)] \right) \prod_{l=0}^{\tau-1} \left( \prod_{j} [1 - p_{j}(k,l)] \right)\\
&\approx p_{i}(k,\tau) \exp \left( -\sum_{j} \int_{0}^{\tau} p_{j}(k,l) \mathrm{d}l \right),
\end{aligned}
\end{equation}
i.e., the waiting time for an inhomogeneous exponential random variable.
The approximation assumes the $p_i(k,\tau)$ are small enough such that at most one event happens each generation, with small probability, and the changes in the $p_i(k,\tau)$ from one generation to the next are small enough that we can approximate a sum across $\tau$ with an integral.
As a technical aside, to speed computation we analytically solve the integrals in Equation \ref{eq:Pikt} under the assumption of unbounded population growth (Equation \ref{eq:ptnt}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Pairwise diversity}
\label{sec:rescue_heterozygosity}

One classic signature of a selective sweep is a dip in genetic diversity around the selected site \citep{smith1974hitch,kaplan1989hitchhiking}. 
Here we consider the average number of nucleotide differences between two randomly sampled sequences, $\pi$ \citep{tajima1983evolutionary}, focusing on sequences of length 1 (i.e., heterozygosity).

We first consider our expectation for $\pi$ at a site that is far enough away from the selected site to be unaffected by hitchhiking, which will provide us with an expectation for  genome-wide average diversity.
While not directly impacted by the sweep, these loci are indirectly impacted as the sweep dictates the severity of the population bottleneck.
Our expectation for $\pi$ at such a site in a population of constant effective size, $N_e$, is simply $\mathbb{E}[\pi | \mathrm{unlinked}] = \theta \equiv 4 N_e U$ \citep{watterson1975number}, with $U$ the per base per generation mutation rate at neutral loci.
Ignoring neutral mutation input during the sweep, the $\pi$ at a sufficiently loosely linked site in a population of changing size is this neutral expectation times the probability a sample of size two does not coalesce during the sweep (c.f., equation 4 in \citealp{slatkin1991pairwise} and equation 7 in \citealp{griffiths1994sampling}),
\begin{equation}\label{eq:EH_rescue_loose}
\mathbb{E}[\pi | \mathrm{unlinked}]\approx \theta \exp \left( - \int_0^{t_f} \frac{1}{2N_e'(\tau)} \mathrm{d}\tau \right). 
\end{equation}
%In a population of constant size we simply use $\mathbb{E}[\pi | \mathrm{unlinked}]=\theta$, which is more accurate as it incorporates neutral mutations occurring during the sweep.

We next consider sites that are close enough to the selected locus that they are directly affected by the selective sweep through hitchhiking.
To keep the analysis simple, we assume that if recombination or mutation moves one of the sampled alleles to the ancestral background before the two samples coalesce then it is as if both samples were on the ancestral background from the start and therefore coalesce with each other as if they were at an unlinked locus (Equation \ref{eq:EH_rescue_loose}).
This assumption ignores the time it takes for the two samples to arrive on the ancestral background and is therefore expected to underestimate diversity at moderately-linked sites.
There is also the possibility that no events occur in the history of the sample during the sweep.
Then, if the sweep arose from mutation or migration or from a single copy of the beneficial allele ($\kappa=1$), the sample must coalesce. 
If we instead start with more than one beneficial copy, $\kappa>1$, it is possible that the two samples had ancestors that were linked to distinct copies of the beneficial allele within the standing variation.
While the coalescent naturally tells us the probability that two lineages do not coalesce during the sweep \citep[since distinct copies of the beneficial allele are exchangeable; e.g.,][]{tavare1984line}, our continuous-time approximation of the coalescent (Equation \ref{eq:Pikt}) implicitly assumes a large number of copies of the beneficial allele at any one time.
To account for the fact that $\kappa$ is finite, and often small, we assume that if under the continuous-time approximation of the coalescent there remain two lineages on the beneficial background at the beginning of the sweep then they coalesce with probability $1/\kappa$ \citep[see the `instantaneous' approximation in][for a similar approach]{anderson2007estimation}.  
Finally, we assume that each copy of the beneficial allele in the standing variance has a recent and unique mutational origin \citep[implying the beneficial allele was sufficiently deleterious before the environmental change; c.f.,][]{prezeworski2005signature,berg2015coalescent}, so that two distinct ancestral lineages are independent draws from a neutral population.
Ignoring neutral mutations during the sweep, a simple approximation for diversity at any location in the genome is then
\begin{equation}\label{eq:EH_rescue}
\begin{aligned}
\mathbb{E}[\pi ] \approx P_\mathrm{off}(2,t_f) \mathbb{E}[\pi | \mathrm{unlinked}] + P_\emptyset(2,t_f) (1 - 1/\kappa) \theta,
\end{aligned}
\end{equation}
where $P_\mathrm{off}(k,t_f) = \int_0^{t_f} [P_\mathrm{rec}(k,\tau) + P_\mathrm{mut}(k,\tau)] \mathrm{d}\tau$ is the probability of recombination or mutation before coalescence during the sweep and $P_\emptyset(k, t_f) = \exp \left( -\sum_{j} \int_{0}^{t_f} p_{j}(k,\tau) \mathrm{d}\tau \right)$ is the probability that no events have occurred in the history of the sample during the sweep ($j\in\{\mathrm{rec},\mathrm{mut},\mathrm{coal}\}$).
Equation \ref{eq:EH_rescue} is a function of recombination distance, $r$, and approaches $\mathbb{E}[\pi | \mathrm{unlinked}]$ as $r$ gets large.

@@

Code used to derive and plot all results (Python scripts and Mathematica notebook) is available at \url{https://github.com/mmosmond/rescueCoalescent.git}.
File S1 refers to the Mathematica notebook, which is also provided in freely accessible forms (e.g., PDF).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{A simultaneous, but independent, sweep and bottleneck}
\label{sec:sweep+bottleneck}
